Overlapping Batches for the Assessment of Solution Quality in Stochastic Programs

### 2013-10-24 omrp.tex
Changed the ref style

\usepackage{latexsym}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{enumerate}
%\usepackage{mathptmx}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsbsy}
\usepackage{amsthm}
\usepackage[pdftex,colorlinks=true,urlcolor=blue,citecolor=black,anchorcolor=black,linkcolor=black]{hyperref}
\usepackage{setspace}

@@

Overlapping Batch Means (OBM) method is widely used in simulation to obtain variance estimators with lower variances. 
In this paper we show that the same benefits can be obtained from overlapping batches when used in a stochastic optimization setting.  
A natural setting where batch means are used in stochastic optimization (esp. in stochastic programming) is for assessing solution quality. 
Assessing solution quality in stochastic programming typically involves generating point and interval estimators of the optimality gap of a given candidate solution using Monte Carlo simulation. 
Because the optimality gap estimators can be difficult to analyze statistically, a common approach is to use nonovelapping batches of data to generate a number of optimality gap estimators. 
These individual estimators are then used to form point and interval estimators using the classical nonoverlapping batches method.
This is referred to as the Multiple Replications Procedure (MRP).   
In this paper, we formulate an ovelapping batches version of MRP, allowing for partial overlap, and provide conditions under which the same variance reduction results can be achieved by overlapping like in the classical setting. 
We also show that the resulting point estimators are strongly consistent and the interval estimators are asymptotically valid. 
We present computational results to examine small-sample behavior. 
Our experiments indicate that the asymptotic variance reduction is achieved with small sample sizes, while bias and coverage probability are unaffected.

@@

Assessing solution quality is important in practice since (\ref{eq:sto_prog}) typically cannot be solved exactly, so one only has an approximate solution $\xh$ without verification of its quality.
Assessing solution quality is also a critical component of stopping criteria in algorithms.
In many fields of optimization, it is common to evaluate an upper bound on the optimality gap through relaxations such as integrality or Lagrangian relations.
In stochastic programming, an approximation of \ref{eq:sto_prog} is often generated by sampling from the distribution of $\xi$.
Let $\xit^1, \xit^2, \dots, \xit^m$ be an independent and identically distributed (iid) sample from the distribution of $\xit$.
A sampling approximation of (\ref{eq:sto_prog}) is given by
\begin{align} \tag{SP$_m$} \label{eq:sto_prog_m}
	z_m^* & = \min_{\x \in X} \frac{1}{m} \sum_{i=1}^m f(\x,\xiti).
\end{align}
We denote an optimal solution to (\ref{eq:sto_prog}) as $\xs$ and an optimal solution to (\ref{eq:sto_prog_m}) as $\xs_m$.

This motivates estimating the upper bound on the optimality gap by using Monte Carlo sampling.
Given a sample size $m$, an upper bound on the optimality gap can be obtained by $\e{f(\xh,\xit)} - \zs \leq \e{f(\xh,\xit)} - \e{\zs_m}$ due to the inequality $\e{\zs_m} \leq \zs$.
This bound improves as the sample size increases, that is, $\e{\zs_m} \leq \e{\zs_{m+1}} \leq \zs$ \cite{Mak1999,norkin_pflug_ruszczynski_98}.  
A straightforward estimate of $\e{f(\xh,\xit)}$ is the sample mean, $\frac{1}{m} \sum_{i=1}^m f(\xh,\xiti)$. 
Instead of $\e{\zs_m}$, simply $\zs_m$ can be used.  
The resulting point estimator is  $G_m(\xh) = \frac{1}{m} \sum_{i=1}^m f(\xh,\xiti) - \zs_m$.  
Here, we assume the same observations $\xit^1, \dots, \xit^m$ are used in both terms of $G_m(\xh)$.

Computing $G_m(\xh)$ involves solving an optimization problem (\ref{eq:sto_prog_m}) to obtain a lower bound estimator $\zs_m$, which complicates the statistical analysis. 
To enable statistical inference, the \emph{Multiple Replications Procedure} (MRP) of Mak et.\ al.\ \cite{Mak1999} generates $k$ independent estimators of $G_m(\xh)$, each using sample size $m$ and averages them to obtain a point estimator.  
To form confidence intervals, the sample variance of these estimators is calculated.  
This is essentially a nonoverlapping batch means type estimator ($k$ batches of size $m$) commonly used in simulation \cite{law_07}. 
We further review MRP estimators in \S \ref{ssec:mrp}.

In this paper, we extend MRP by overlapping the batches similar to the overlapping batch means (OBM) introduced by Meketon and Schmeiser \cite{Meketon1984}. 
As an example, suppose each batch contains $m$ observations and that there are a total of $2m$ observations.
In the nonoverlapping case, there are only two batches: the first batch contains observations $\xit^1, \xit^2, \dots, \xit^m$ and the second batch contains observations $\xit^{m+1}, \xit^{m+2}, \dots, \xit^{2m}$.   
Meketon and Schmeiser \cite{Meketon1984} overlap the batches so that the first batch is as before, the second batch now consists of $\xit^2, \xit^3, \dots, \xit^{m+1}$, the third batch contains observations $\xit^3, \xit^4, \dots, \xit^{m+2}$ and so on. 
As a result, the corresponding batch means are no longer independent. 
However, by simply reusing the data in this fashion, and seemingly in a contradictory way that induces dependence, one obtains a better variance estimator: a variance estimator that asymptotically has two-thirds of the variance of the usual batch means variance estimator but the bias of the two estimators is approximately the same.  
The idea of overlapping batches can be used for variance estimators of non-means \cite{SAH90} and recently, has been extended to estimators based on standardized time series (e.g., area, Cram\'{e}r-von Mises) for steady-state simulation output analysis \cite{Alexopoulos01012007,Alexopoulos2007}.

We explore the use of overlapping batches for assessing solution quality in stochastic programs.
Similar to Welch \cite{Welch1987} and Song and Schmeiser \cite{Song1992}, we consider partial overlap between batches to reduce computational effort in our setting.
The contributions of this paper are:
\begin{itemize}
	\item Formulate the overlapping multiple replications procedure for stochastic programs.
	\item Give conditions under which the point estimators are strongly consistent (that is, they converge almost surely (a.s.) as opposed in to in probability; referred to simply as consistent from this point on).
	\item Demonstrate that the classical variance reduction results are achieved for stochastic programs.
	\item Show the asymptotic validity of the resulting confidence intervals.
	\item Present computational results on two-stage stochastic linear and integer programs with recourse to examine small sample behavior.  
		Our experiments indicate that the asymptotic variance reduction is achieved with small sample sizes, while bias and coverage probability are unaffected.
\end{itemize}

@@

\subsection{Overlapping Batch Means} \label{ssec:obm}
Consider a covariance stationary stochastic process which has mean $\mu$ and variance $\sigma^2$.  
The task is to estimate $\mu$ given some realization of the stochastic process $\y = (y^1, y^2, \dots, y^n)$.  
Typically, this involves forming a confidence interval of the form $[L_\alpha(\y), U_\alpha(\y)]$ for a given level of significance $\alpha$ such that $\p{L_\alpha(\y) \leq \mu \leq U_\alpha(\y)} = 1 - \alpha$.  
The usual estimator for $\mu$ is the sample mean $\ybb = \frac{1}{n} \sum_{i=1}^n y^i$, which is an unbiased estimator even in the presence of correlated data.  
However, the usual variance estimator $\frac{1}{n-1} \sum_{i=1}^n (y^i - \ybb)^2$ can be inappropriate to use in the presence of correlated data, resulting in inaccurate interval estimators.  
To overcome this difficulty, various variance estimators have been proposed in the simulation literature; see, e.g., \cite{law_07}.  
We briefly review two of these variance estimators: nonoverlapping and overlapping batch means.

@@

Overlapping batch means modifies this idea by taking batches given by $\yb_j(m) = \frac{1}{m} \sum_{i=1}^{m} y^{(j-1)+i}$, for $j = 1, \dots, n-m+1$.  
This is what we call the \emph{maximally-overlapping} batch means method, because $m-1$ observations are common to---and only one observation changes between---adjacent batches.  
Given this, the sample variance estimator is updated to
\begin{equation} \label{eq:overlap_variance_def}
	\vart{\ybb} = \frac{1}{(\tfrac{n}{m}-1)(n - m + 1)}\sum_{j=1}^{n-m+1} (\yb_j(m) - \ybb)^2.
\end{equation}
Meketon and Schmeiser \cite{Meketon1984} show several attractive properties of the overlapping batches variance estimator, the two most important of which are: (i) the overlapping variance estimator has nearly the same bias as the standard nonoverlapping variance estimator, and (ii) the overlapping estimator has only two-thirds of the asymptotic variance of the nonoverlapping one.  
That is, 
\begin{equation} \label{eq:asym_var}
	\frac{ \var{\vart{\ybb}} }{ \var{\varhat(\ybb)} } \rightarrow \frac{2}{3}
\end{equation}
in the limit as batch size $m$ and then the number of batches $(n/m)$ tend to infinity.  
Note that the degrees of freedom in (\ref{eq:overlap_variance_def}) is slightly different than in \cite{Meketon1984}.  
The degrees of freedom in (\ref{eq:overlap_variance_def}) makes $\vart{\ybb}$ an unbiased estimator for iid data for all finite $m$ and $n$ with the same asymptotic benefits \cite{Song1992}.  
The $(1-\alpha)$-level approximate CI is similarly formed by $\ybb \pm t_{3(k-1)/2,\alpha/2} \sqrt{\vart{\ybb}}$, with an adjustment to the degrees of freedom.

Welch \cite{Welch1987} established the relationship between OBM and spectral estimators and also considered partial overlap, see also \cite{Song1992}.  
In our application to assessment of solution quality, we also consider varying the amount of overlap between neighboring batches.  
We defer this discussion on variable overlap to \S \ref{ssec:overlap} and continue with a brief review of MRP.

@@

Notice that the minimization in the second term on the right-hand side of (\ref{eq:mrp_gap}) gives rise to a biased gap estimator (recall the upper bound on the optimality gap).  
Thus, we can expect that the true probability of the optimality gap residing within the confidence interval to be greater than the $1 - \alpha$ suggested by the above calculation.  
This is shown empirically in Bayraksan and Morton \cite{Bayraksan2006}, which also suggests additional methods for using a smaller number of replications (e.g., 1 or 2) with an alternative variance estimator to compute a confidence interval.  
See also \cite{bayraksan_morton_09,partani2006jackknife,partani_07} for variations of MRP aimed to reduce bias and variance.

An advantage of MRP is its applicability to a wide range of problems.  
With iid sampling, (\ref{eq:sto_prog}) can be linear or nonlinear, $\X$ can include integrality constraints or not.  
It is also easy to implement, thus, has been applied to a variety of problems, see e.g., \cite{bertocchi_etal_99,janjarassuk_linderoth_08,santoso_ahmed_etal_05}.  
Recently, the approach of (nonoverlapping) batching has been used for assessing solution quality of stochastic programs with finitely many expected value \cite{wang_ahmed_08} and stochastic dominance \cite{hu2010sample} constraints.

@@

As before, let $m$ denote the batch size, $n$ the total sample size and $k = \lfloor \tfrac{n}{m} \rfloor$ be the number of nonoverlapping batches.  
In this paper, we use the {\it batch nonoverlap parameter} $1 \leq \gamma \leq m$ to denote how much neighboring batches do {\it not} overlap.  
For instance, $\gamma = m$ corresponds to the classical case of nonoverlapping batches and $\gamma = 1$ corresponds to the maximally overlapping case of Meketon and Schmeiser \cite{Meketon1984}.  
The sample mean of each batch estimator is calculated similarly, 
$
\yb_j = \frac{1}{m} \sum_{i=1}^m y^{\gamma(j-1) + i},\ \  j = 1, 2, \dots, \nb, 
$
where $\nb$ is the number of batches used given $n$, $m$ and $\gamma$.  
The sample variance estimator given in (\ref{eq:overlap_variance_def}) is changed to
\begin{equation} \label{eq:variance_gamma}
	\vartg{\ybb} = \frac{1}{\left( \tfrac{n}{m} - 1 \right) \left( \nb \right)}  \sum_{j=1}^{\nb} (\yb_j - \ybb)^2.
\end{equation}
Note that with $\gamma=1$, (\ref{eq:variance_gamma}) reduces to (\ref{eq:overlap_variance_def}) and with $\gamma=m$ and $n=mk$, (\ref{eq:variance_gamma}) reduces to (\ref{eq:var}).

The amount of asymptotic variance reduction in this estimator depends on the asymptotic ratio of $\gamma/m$, which we denote by $\gammab$.  
For example, when $\gamma = 1$ (or $\gammab=0$; the maximally overlapping case), the variance is reduced to two-thirds ($66.67\%$) of the original nonoverlapping case, as given in (\ref{eq:asym_var}).  
When 75\% of observations overlap (i.e., when $\gamma = m/4$ or $\gammab = 1/4$), the variance is $33/48$th of the original ($68.75\%$), which is near-optimal.   
When only half of the observations overlap ($\gammab = 1/2$) the variance is 75\% of original.  
In general, from the spectral analysis given in Welch \cite{Welch1987}, the variance is reduced to
\begin{equation} \label{eq:var_reduct_formula}
	\gammab \left( 1 + 2 \sum_{j=1}^{\cogb-1} (1-j\gammab)^2 \right) = -\gammab\left( 1 - 2\gammab\cogb + \gammab^2\cogb^2 \right) + 2\gammab\cogb\left(1-\gammab\cogb\right) + \frac{2}{3}\gammab^3\cogb^3 + \frac{1}{3}\gammab^3\cogb
\end{equation}
of the nonoverlapping case.  
If $\ogb$ is an integer, the terms in parentheses vanish and (\ref{eq:var_reduct_formula}) simplifies to $\frac{2+\gammab^2}{3}$.

A $(1-\alpha)$-level approximate CI on the mean can be formed using variably overlapping batches by $\ybb \pm t_{d_{\gamma}(k-1),\alpha/2}$ $\sqrt{\vartg{\ybb}}$, where the degrees of freedom increase $d_{\gamma}$ is the multiplicative inverse of (\ref{eq:var_reduct_formula}) \cite{Welch1987}.  
If $\ogb$ is an integer, $d_{\gamma} = \frac{3}{2+\gammab^2}$ and for maximally overlapping batches, $d_{\gamma} = \frac{3}{2}$.

@@

The results on overlapping batch means occur in the limit as $n, m, k=n/m \rightarrow \infty$ \cite{damerdji1994strong,damerdji1995mean,Meketon1984,Song1992,Welch1987}.  
Let $n_l, m_l$ and $k_l$ be sequences of numbers satisfying these requirements, then the limits will be taken as $l \rightarrow \infty$.  
We assume a batching structure where $m_l = \lfloor n_l ^r \rfloor$ for some $0<r<1$ as is typical in simulation, and $n_l$ tends to infinity as $l \rightarrow \infty$.  
In addition, we may desire that the batch nonoverlap parameter change with $l$, to ensure that $\gamma_l / m_l = \gammab_l$ converges to the constant $\gammab$.  
Now we are ready to define the estimators for OMRP.

@@

First, we restrict our attention to iid sampling by A1.
The classical variance-reduction results for overlapping batches show that $\var{n_l VG_l}$ is reduced by overlapping, thus requiring at least finite fourth moments. 
Finally, our proof relies on an exponential rate of convergence of sampled solution to the true solution of the problem.  
Assumption A3 can be satisfied in several ways.  
Shapiro and Homem-de-Mello \cite{shapiro2000rate} establish rates of convergence for a class of (\ref{eq:sto_prog}), which include two-stage stochastic linear programs with recourse when the support of $\xit$ is finite and the optimal solution is unique.   
Kleywegt, Shapiro,and Homem-de-Mello \cite{kleywegt2002sample} show that assumption A3 can be satisfied for discrete stochastic programs (i.e., when $X$ is a finite set) under the condition of a unique optimal solution $\xs$, and that the moment generating function of $f(\xs,\xit) - f(\x,\xit)$ for $x \in X \backslash \{\xs\}$ is finite valued on $\R$.  
This class of problems includes stochastic integer programs with recourse.
  We present computational results on both stochastic linear and integer programs with recourse in \S \ref{sec:comp}.

@@

Strong consistency of the variance estimator $n_l VD_l$ have been studied in \cite{damerdji1994strong} for the cases of nonoverlapping and maximally overlapping batches.  
Under the additional moment assumption
\begin{description}
	\item[A2$^\prime$]  $\exists \epsilon > 0$ such that $\sup_{\x \in X} \e{f(\x,\xi)^{4+\epsilon}} < \infty$,
\end{description}
for a sampling scheme of $m_l = \lfloor n_l^r \rfloor$, Corollaries 3.4 and 4.2 of \cite{damerdji1994strong} show that $n_l VD_l$ is consistent if $\tfrac{2}{4+\epsilon} < r < 1$ for nonoverlapping batches and $\tfrac{2}{4+\epsilon} < r < \tfrac{1}{2}$ for maximally overlapping batches.

@@

Meketon and Schmeiser \cite{Meketon1984} show that $\frac{n_l}{m_l}\var{n_l VD_l} \rightarrow \alpha \sigma^4_{\xh}$ where $\alpha = \frac{4}{3}$ for maximally overlapping batches and $\alpha = 2$ for nonoverlapping batches, resulting in the decrease of variance for the maximally overlapping batches estimator.  
Welch \cite{Welch1987} expands this result to variable amounts of overlap, showing that $\alpha$ is twice the value of (\ref{eq:var_reduct_formula}).  
Damerdji \cite{damerdji1995mean} shows that convergence of $\frac{n_l}{m_l}\var{n_l VD_l}$ to the appropriate $\alpha \sigma^4_{\xh}$, and thus the mean-square consistency of $n_l VD_l$, is guaranteed for $\tfrac{1}{2} + \tfrac{1}{4+\epsilon} < r < 1$ for both nonoverlapping and maximally overlapping batches under the batching structure $m_l = \lfloor n_l^r \rfloor$ and $4+\epsilon$ finite moments assumption A2$^\prime$.  
(With A2, the range of $r$ that ensures mean-square consistency can be found by setting $\epsilon=0$.)

>>>>

\usepackage{latexsym}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{enumerate}
%\usepackage{mathptmx}
\usepackage[square]{natbib}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsbsy}
\usepackage{amsthm}
\usepackage[pdftex,colorlinks=true,urlcolor=blue,citecolor=black,anchorcolor=black,linkcolor=black]{hyperref}
\usepackage{setspace}

@@

Overlapping Batch Means (OBM) method, proposed by \citet{Meketon1984}, is widely used in simulation to obtain variance estimators with lower variances. 
In this paper we show that the same benefits can be obtained from overlapping batches when used in a stochastic optimization setting.  
A natural setting where batch means are used in stochastic optimization (esp. in stochastic programming) is for assessing solution quality. 
Assessing solution quality in stochastic programming typically involves generating point and interval estimators of the optimality gap of a given candidate solution using Monte Carlo simulation. 
Because the optimality gap estimators can be difficult to analyze statistically, a common approach is to use nonovelapping batches of data to generate a number of optimality gap estimators. 
These individual estimators are then used to form point and interval estimators using the classical nonoverlapping batches method.
This is referred to as the Multiple Replications Procedure (MRP) and was initially discussed in \citep{Mak1999}.   
In this paper, we formulate an ovelapping batches version of MRP, allowing for partial overlap, and provide conditions under which the same variance reduction results can be achieved by overlapping like in the classical setting. 
We also show that the resulting point estimators are strongly consistent and the interval estimators are asymptotically valid. 
We present computational results to examine small-sample behavior. 
Our experiments indicate that the asymptotic variance reduction is achieved with small sample sizes, while bias and coverage probability are unaffected.

@@

Assessing solution quality is important in practice since (\ref{eq:sto_prog}) typically cannot be solved exactly, so one only has an approximate solution $\xh$ without verification of its quality.
Assessing solution quality is also a critical component of stopping criteria in algorithms.

In stochastic programming, an approximation of \ref{eq:sto_prog} is often generated by sampling from the distribution of $\xi$.
Let $\xit^1, \xit^2, \dots, \xit^m$ be an independent and identically distributed (iid) sample from the distribution of $\xit$.
A sampling approximation of (\ref{eq:sto_prog}) is given by
\begin{align} \tag{SP$_m$} \label{eq:sto_prog_m}
	z_m^* & = \min_{\x \in X} \frac{1}{m} \sum_{i=1}^m f(\x,\xiti).
\end{align}
We denote an optimal solution to (\ref{eq:sto_prog}) as $\xs$ and an optimal solution to (\ref{eq:sto_prog_m}) as $\xs_m$.

In many fields of optimization, it is common to evaluate an upper bound on the optimality gap through relaxations such as integrality or Lagrangian relations.

This motivates estimating the upper bound on the optimality gap by using Monte Carlo sampling.
Given a sample size $m$, an upper bound on the optimality gap can be obtained by $\e{f(\xh,\xit)} - \zs \leq \e{f(\xh,\xit)} - \e{\zs_m}$ due to the inequality $\e{\zs_m} \leq \zs$.
This bound improves as the sample size increases, that is, $\e{\zs_m} \leq \e{\zs_{m+1}} \leq \zs$ \citep{Mak1999,norkin_pflug_ruszczynski_98}.  
A straightforward estimate of $\e{f(\xh,\xit)}$ is the sample mean, $\frac{1}{m} \sum_{i=1}^m f(\xh,\xiti)$. 
Instead of $\e{\zs_m}$, simply $\zs_m$ can be used.  
The resulting point estimator is  $G_m(\xh) = \frac{1}{m} \sum_{i=1}^m f(\xh,\xiti) - \zs_m$.  
Here, we assume the same observations $\xit^1, \dots, \xit^m$ are used in both terms of $G_m(\xh)$.

Computing $G_m(\xh)$ involves solving an optimization problem (\ref{eq:sto_prog_m}) to obtain a lower bound estimator $\zs_m$, which complicates the statistical analysis. 
To enable statistical inference, the \emph{Multiple Replications Procedure} (MRP) of Mak et.\ al.\ \citep{Mak1999} generates $k$ independent estimators of $G_m(\xh)$, each using sample size $m$ and averages them to obtain a point estimator.  
To form confidence intervals, the sample variance of these estimators is calculated.  
This is essentially a nonoverlapping batch means type estimator ($k$ batches of size $m$) commonly used in simulation \citep{law_07}. 
We further review MRP estimators in \S \ref{ssec:mrp}.

In this paper, we extend MRP by overlapping the batches similar to the overlapping batch means (OBM) introduced by Meketon and Schmeiser \citep{Meketon1984}. 
As an example, suppose each batch contains $m$ observations and that there are a total of $2m$ observations.
In the nonoverlapping case, there are only two batches: the first batch contains observations $\xit^1, \xit^2, \dots, \xit^m$ and the second batch contains observations $\xit^{m+1}, \xit^{m+2}, \dots, \xit^{2m}$.   
Meketon and Schmeiser \citep{Meketon1984} overlap the batches so that the first batch is as before, the second batch now consists of $\xit^2, \xit^3, \dots, \xit^{m+1}$, the third batch contains observations $\xit^3, \xit^4, \dots, \xit^{m+2}$ and so on. 
As a result, the corresponding batch means are no longer independent. 
However, by simply reusing the data in this fashion, and seemingly in a contradictory way that induces dependence, one obtains a better variance estimator: a variance estimator that asymptotically has two-thirds of the variance of the usual batch means variance estimator but the bias of the two estimators is approximately the same.  
The idea of overlapping batches can be used for variance estimators of non-means \citep{SAH90} and recently, has been extended to estimators based on standardized time series (e.g., area, Cram\'{e}r-von Mises) for steady-state simulation output analysis \citep{Alexopoulos01012007,Alexopoulos2007}.

We explore the use of overlapping batches for assessing solution quality in stochastic programs.
Similar to Welch \citep{Welch1987} and Song and Schmeiser \citep{Song1992}, we consider partial overlap between batches to reduce computational effort in our setting.
The contributions of this paper are:
\begin{itemize}
	\item Formulate the overlapping multiple replications procedure for stochastic programs.
	\item Give conditions under which the point estimators are strongly consistent (that is, they converge almost surely (a.s.) as opposed in to in probability; referred to simply as consistent from this point on).
	\item Demonstrate that the classical variance reduction results are achieved for stochastic programs.
	\item Show the asymptotic validity of the resulting confidence intervals.
	\item Present computational results on two-stage stochastic linear and integer programs with recourse to examine small sample behavior.  
		Our experiments indicate that the asymptotic variance reduction is achieved with small sample sizes, while bias and coverage probability are unaffected.
\end{itemize}

@@

\subsection{Overlapping Batch Means} \label{ssec:obm}
Consider a covariance stationary stochastic process which has mean $\mu$ and variance $\sigma^2$.  
The task is to estimate $\mu$ given some realization of the stochastic process $\y = (y^1, y^2, \dots, y^n)$.  
Typically, this involves forming a confidence interval of the form $[L_\alpha(\y), U_\alpha(\y)]$ for a given level of significance $\alpha$ such that $\p{L_\alpha(\y) \leq \mu \leq U_\alpha(\y)} = 1 - \alpha$.  
The usual estimator for $\mu$ is the sample mean $\ybb = \frac{1}{n} \sum_{i=1}^n y^i$, which is an unbiased estimator even in the presence of correlated data.  
However, the usual variance estimator $\frac{1}{n-1} \sum_{i=1}^n (y^i - \ybb)^2$ can be inappropriate to use in the presence of correlated data, resulting in inaccurate interval estimators.  
To overcome this difficulty, various variance estimators have been proposed in the simulation literature; see, e.g., \citep{law_07}.  
We briefly review two of these variance estimators: nonoverlapping and overlapping batch means.

@@

Overlapping batch means modifies this idea by taking batches given by $\yb_j(m) = \frac{1}{m} \sum_{i=1}^{m} y^{(j-1)+i}$, for $j = 1, \dots, n-m+1$.  
This is what we call the \emph{maximally-overlapping} batch means method, because $m-1$ observations are common to---and only one observation changes between---adjacent batches.  
Given this, the sample variance estimator is updated to
\begin{equation} \label{eq:overlap_variance_def}
	\vart{\ybb} = \frac{1}{(\tfrac{n}{m}-1)(n - m + 1)}\sum_{j=1}^{n-m+1} (\yb_j(m) - \ybb)^2.
\end{equation}
Meketon and Schmeiser \citep{Meketon1984} show several attractive properties of the overlapping batches variance estimator, the two most important of which are: (i) the overlapping variance estimator has nearly the same bias as the standard nonoverlapping variance estimator, and (ii) the overlapping estimator has only two-thirds of the asymptotic variance of the nonoverlapping one.  
That is, 
\begin{equation} \label{eq:asym_var}
	\frac{ \var{\vart{\ybb}} }{ \var{\varhat(\ybb)} } \rightarrow \frac{2}{3}
\end{equation}
in the limit as batch size $m$ and then the number of batches $(n/m)$ tend to infinity.  
Note that the degrees of freedom in (\ref{eq:overlap_variance_def}) is slightly different than in \citep{Meketon1984}.  
The degrees of freedom in (\ref{eq:overlap_variance_def}) makes $\vart{\ybb}$ an unbiased estimator for iid data for all finite $m$ and $n$ with the same asymptotic benefits \citep{Song1992}.  
The $(1-\alpha)$-level approximate CI is similarly formed by $\ybb \pm t_{3(k-1)/2,\alpha/2} \sqrt{\vart{\ybb}}$, with an adjustment to the degrees of freedom.

Welch \citep{Welch1987} established the relationship between OBM and spectral estimators and also considered partial overlap, see also \citep{Song1992}.  
In our application to assessment of solution quality, we also consider varying the amount of overlap between neighboring batches.  
We defer this discussion on variable overlap to \S \ref{ssec:overlap} and continue with a brief review of MRP.

@@

Notice that the minimization in the second term on the right-hand side of (\ref{eq:mrp_gap}) gives rise to a biased gap estimator (recall the upper bound on the optimality gap).  
Thus, we can expect that the true probability of the optimality gap residing within the confidence interval to be greater than the $1 - \alpha$ suggested by the above calculation.  
This is shown empirically in Bayraksan and Morton \citep{Bayraksan2006}, which also suggests additional methods for using a smaller number of replications (e.g., 1 or 2) with an alternative variance estimator to compute a confidence interval.  
See also \citep{bayraksan_morton_09,partani2006jackknife,partani_07} for variations of MRP aimed to reduce bias and variance.

An advantage of MRP is its applicability to a wide range of problems.  
With iid sampling, (\ref{eq:sto_prog}) can be linear or nonlinear, $\X$ can include integrality constraints or not.  
It is also easy to implement, thus, has been applied to a variety of problems, see e.g., \citep{bertocchi_etal_99,janjarassuk_linderoth_08,santoso_ahmed_etal_05}.  
Recently, the approach of (nonoverlapping) batching has been used for assessing solution quality of stochastic programs with finitely many expected value \citep{wang_ahmed_08} and stochastic dominance \citep{hu2010sample} constraints.

@@

As before, let $m$ denote the batch size, $n$ the total sample size and $k = \lfloor \tfrac{n}{m} \rfloor$ be the number of nonoverlapping batches.  
In this paper, we use the {\it batch nonoverlap parameter} $1 \leq \gamma \leq m$ to denote how much neighboring batches do {\it not} overlap.  
For instance, $\gamma = m$ corresponds to the classical case of nonoverlapping batches and $\gamma = 1$ corresponds to the maximally overlapping case of Meketon and Schmeiser \citep{Meketon1984}.  
The sample mean of each batch estimator is calculated similarly, 
$
\yb_j = \frac{1}{m} \sum_{i=1}^m y^{\gamma(j-1) + i},\ \  j = 1, 2, \dots, \nb, 
$
where $\nb$ is the number of batches used given $n$, $m$ and $\gamma$.  
The sample variance estimator given in (\ref{eq:overlap_variance_def}) is changed to
\begin{equation} \label{eq:variance_gamma}
	\vartg{\ybb} = \frac{1}{\left( \tfrac{n}{m} - 1 \right) \left( \nb \right)}  \sum_{j=1}^{\nb} (\yb_j - \ybb)^2.
\end{equation}
Note that with $\gamma=1$, (\ref{eq:variance_gamma}) reduces to (\ref{eq:overlap_variance_def}) and with $\gamma=m$ and $n=mk$, (\ref{eq:variance_gamma}) reduces to (\ref{eq:var}).

The amount of asymptotic variance reduction in this estimator depends on the asymptotic ratio of $\gamma/m$, which we denote by $\gammab$.  
For example, when $\gamma = 1$ (or $\gammab=0$; the maximally overlapping case), the variance is reduced to two-thirds ($66.67\%$) of the original nonoverlapping case, as given in (\ref{eq:asym_var}).  
When 75\% of observations overlap (i.e., when $\gamma = m/4$ or $\gammab = 1/4$), the variance is $33/48$th of the original ($68.75\%$), which is near-optimal.   
When only half of the observations overlap ($\gammab = 1/2$) the variance is 75\% of original.  
In general, from the spectral analysis given in Welch \citep{Welch1987}, the variance is reduced to
\begin{equation} \label{eq:var_reduct_formula}
	\gammab \left( 1 + 2 \sum_{j=1}^{\cogb-1} (1-j\gammab)^2 \right) = -\gammab\left( 1 - 2\gammab\cogb + \gammab^2\cogb^2 \right) + 2\gammab\cogb\left(1-\gammab\cogb\right) + \frac{2}{3}\gammab^3\cogb^3 + \frac{1}{3}\gammab^3\cogb
\end{equation}
of the nonoverlapping case.  
If $\ogb$ is an integer, the terms in parentheses vanish and (\ref{eq:var_reduct_formula}) simplifies to $\frac{2+\gammab^2}{3}$.

A $(1-\alpha)$-level approximate CI on the mean can be formed using variably overlapping batches by $\ybb \pm t_{d_{\gamma}(k-1),\alpha/2}$ $\sqrt{\vartg{\ybb}}$, where the degrees of freedom increase $d_{\gamma}$ is the multiplicative inverse of (\ref{eq:var_reduct_formula}) \citep{Welch1987}.  
If $\ogb$ is an integer, $d_{\gamma} = \frac{3}{2+\gammab^2}$ and for maximally overlapping batches, $d_{\gamma} = \frac{3}{2}$.

@@

The results on overlapping batch means occur in the limit as $n, m, k=n/m \rightarrow \infty$ \citep{damerdji1994strong,damerdji1995mean,Meketon1984,Song1992,Welch1987}.  
Let $n_l, m_l$ and $k_l$ be sequences of numbers satisfying these requirements, then the limits will be taken as $l \rightarrow \infty$.  
We assume a batching structure where $m_l = \lfloor n_l ^r \rfloor$ for some $0<r<1$ as is typical in simulation, and $n_l$ tends to infinity as $l \rightarrow \infty$.  
In addition, we may desire that the batch nonoverlap parameter change with $l$, to ensure that $\gamma_l / m_l = \gammab_l$ converges to the constant $\gammab$.  
Now we are ready to define the estimators for OMRP.

@@

First, we restrict our attention to iid sampling by A1.
The classical variance-reduction results for overlapping batches show that $\var{n_l VG_l}$ is reduced by overlapping, thus requiring at least finite fourth moments. 
Finally, our proof relies on an exponential rate of convergence of sampled solution to the true solution of the problem.  
Assumption A3 can be satisfied in several ways.  
Shapiro and Homem-de-Mello \citep{shapiro2000rate} establish rates of convergence for a class of (\ref{eq:sto_prog}), which include two-stage stochastic linear programs with recourse when the support of $\xit$ is finite and the optimal solution is unique.   
Kleywegt, Shapiro,and Homem-de-Mello \citep{kleywegt2002sample} show that assumption A3 can be satisfied for discrete stochastic programs (i.e., when $X$ is a finite set) under the condition of a unique optimal solution $\xs$, and that the moment generating function of $f(\xs,\xit) - f(\x,\xit)$ for $x \in X \backslash \{\xs\}$ is finite valued on $\R$.  
This class of problems includes stochastic integer programs with recourse.
  We present computational results on both stochastic linear and integer programs with recourse in \S \ref{sec:comp}.

@@

Strong consistency of the variance estimator $n_l VD_l$ have been studied in \citep{damerdji1994strong} for the cases of nonoverlapping and maximally overlapping batches.  
Under the additional moment assumption
\begin{description}
	\item[A2$^\prime$]  $\exists \epsilon > 0$ such that $\sup_{\x \in X} \e{f(\x,\xi)^{4+\epsilon}} < \infty$,
\end{description}
for a sampling scheme of $m_l = \lfloor n_l^r \rfloor$, Corollaries 3.4 and 4.2 of \citep{damerdji1994strong} show that $n_l VD_l$ is consistent if $\tfrac{2}{4+\epsilon} < r < 1$ for nonoverlapping batches and $\tfrac{2}{4+\epsilon} < r < \tfrac{1}{2}$ for maximally overlapping batches.

@@

Meketon and Schmeiser \citep{Meketon1984} show that $\frac{n_l}{m_l}\var{n_l VD_l} \rightarrow \alpha \sigma^4_{\xh}$ where $\alpha = \frac{4}{3}$ for maximally overlapping batches and $\alpha = 2$ for nonoverlapping batches, resulting in the decrease of variance for the maximally overlapping batches estimator.  
Welch \citep{Welch1987} expands this result to variable amounts of overlap, showing that $\alpha$ is twice the value of (\ref{eq:var_reduct_formula}).  
Damerdji \citep{damerdji1995mean} shows that convergence of $\frac{n_l}{m_l}\var{n_l VD_l}$ to the appropriate $\alpha \sigma^4_{\xh}$, and thus the mean-square consistency of $n_l VD_l$, is guaranteed for $\tfrac{1}{2} + \tfrac{1}{4+\epsilon} < r < 1$ for both nonoverlapping and maximally overlapping batches under the batching structure $m_l = \lfloor n_l^r \rfloor$ and $4+\epsilon$ finite moments assumption A2$^\prime$.  
(With A2, the range of $r$ that ensures mean-square consistency can be found by setting $\epsilon=0$.)
